---
title: "Simulating Sampling"
author: "Rachel Saidi"
format: html
editor: visual
---

## Simulating Sampling

In this activity, we will use simulation on a pre-built dataset to explore **The Central Limit Theorem**

### Load the data, "storms"

```{r warning=FALSE, message=FALSE}
library(dplyr)
data("storms")
?storms  # documentation about variables
```

### Name the variable for wind

Since we are just exploring this single quantitative varible, we will bypass needing to require **storms\$**

```{r}
wind <- storms$wind
mean(wind)
sd(wind)
```

### View a histogram of the population data

```{r}
hist(wind, main = "Maximum sustained Wind Speed (in knots)")
```

```{r}
# What is the shape of the population distribution of wind?
```

### Create a sample of the data using **sample**

We will sample **with replacement**

```{r}
sample(wind, size = 100, replace = TRUE)
mean(sample(wind, size = 100, replace = TRUE))
sd(sample(wind, size = 100, replace = TRUE))
```

### View the histogram of the sample distribution of wind

```{r}
hist(sample(wind, size = 100, replace = TRUE), main = "Sample of Max Wind Speed (in knots)")
```

```{r}
# What is the shape of the sample distribution of wind?

# What does this histogram show that is different than the population plot?
```

### From the sample, create a **SAMPLING** distribution

```{r}
r <- replicate(1000, mean(sample(wind, size=100, replace=TRUE)))
mean(r)
sd(r)
```

### View the histogram of the **sampling distribution**

```{r}
hist(r, main = "Sampling Distribution of Wind Speed (in knots)")
```

```{r}
# What is the shape of the sampling distribution of wind? 

# What does this histogram show that is different than the sample plot?
```
