---
title: "chapter 13-14 statistical inference"
author: "Rachel Saidi"
format: html
editor: visual
---

## Chapter 13 - Mathematical Modeling with Normal Distribution Calculations

### Example2:

Head lengths of brushtail possums follow a nearly normal distribution with mean 92.6 mm and standard deviation 3.6 mm.

a.  Compute the Z scores for possums with head lengths of 95.4 mm and 85.8 mm.

b.  A brushtail possum is found to have a head length of 85.8 mm. What percentile of possums have head lengths 85.8 mm or smaller?


*pnorm(y-observation, mean, sd) - calculate the area under the curve (percentile) up to that y-obersvation*

c.  A brushtail possum is found to have a head length of 95.4 mm. What percentile of possums have head lengths 95.4 mm or more?

d. What is the percentile of possums with head lengths BETWEEN 85.8 and 95.4 mm?

Percentile MEANS cumulative area of the LEFT tail 

```{r}
# Answers 
#a
(95.4-92.6)/3.6
(85.8-92.6)/3.6
#b   pnorm(cut off value of the left tail, mean, sd)
pnorm(85.8, mean = 92.6, sd = 3.6)
#c
1-pnorm(95.4, mean = 92.6, sd = 3.6)
##c (alternative code)
pnorm(95.4, mean = 92.6, sd = 3.6, lower.tail = FALSE)

##d
pnorm(95.4, 92.6, 3.6) - pnorm(85.8, 92.6, 3.6)
```

## To draw the pictures of these areas, use this code:

```{r}
#b drawing
openintro::normTail(m = 92.6, s = 3.6, L = 85.8) 
     # notice L is used to shade the lower tail

#c drawing
openintro::normTail(m = 92.6, s = 3.6, U = 95.4)
      # notice U is used to shade the upper tail
```

### Example3:

Head lengths of brushtail possums follow a nearly normal distribution with mean 92.6 mm and standard deviation 3.6 mm.

a.  What possum head length is in the bottom 20th percentile?

b.  What head length is in the top 5th percentile? 

"top 5th percentile" to "95th percentile"

```{r}
# Answers: (note p must be a percentile, which means it represents area to the left of the cutoff)
#a
qnorm(p=.20, m=92.6, s=3.6) 
#b
qnorm(p=.95, m=92.6, s=3.6)
```

## How to shade for qnorm?

```{r}
# be sure to install library package "asbio" before running this code

library(asbio)
# Shade the lower tail up to your qnorm value
shade.norm(x = qnorm(0.20), tail = "lower")
```



