---
title: "chapter 1 notes"
format: docx
editor: visual
---

## load the libraries

```{r message = FALSE}
library(tidyverse)
library(openintro)
```

## view high school and beyond data class

```{r}
data("loan50")
```

## Scan the first 6 rows of the data

```{r}
head(loan50)
```


## What factors are in the "homeowners" variable?

```{r}
unique(loan50$homeownership)
```
*Notice there are three levels:  "rent", "mortgage", and "own"*

## filter for renters rather than homeowners

```{r}
loan50_rent <- loan50 |>  # |> piping operator - "and then"
  filter(homeownership == "rent")
```

## how many factors for "loan_purpose" are there?

```{r}
unique(loan50$loan_purpose)
```

## What kind of values are in the "loan_amount" variable?

```{r}
summary(loan50$loan_amount)
# now define the mean loan amount
avg_loan <- mean(loan50$loan_amount)
```

*We can see the average loan amount is $17083.*

## Create a new variable (using mutate)

categorize loan_cat as: "at or above average" and "below average" loan amount

```{r}
loan_new <- loan50 |>
  mutate(loan_cat = if_else(loan_amount < avg_loan, "below average","at or above average"))
# view the newly created tibble
head(loan_new)
```

## how many total renters are there?

Notice a new variable, n, is created to show the counts.

```{r}
loan50 |> 
  count(homeownership)
```

```{r}
ggplot(data = loan50, aes(x = loan_amount, y = annual_income, color = homeownership)) +
  geom_point()
```


## Now use "group_by" to do the same thing

Notice if you create a new tibble with the backwards arrow, you have to "print" the tibble by naming it or calling head(new_df)

```{r}
loan_df <- loan50 |> 
  group_by(homeownership) |> 
  count()
loan_df
```

## Calculate proportions of renters between 2 groups under loan_cat

This method (using group_by first and then count) provides marginal row proportions

```{r}
prop1 <- loan_new  |>
  # Group by "group"
  group_by(loan_cat) |>
  count(homeownership) |>
  # Create new variable, prop, using mutate
  mutate(prop = n/sum(n)) 
prop1 
```

Filter further for homeownership == "rent" to compare whether loan_cat had higher proportions of those at or above the mean loan amount.

```{r}

prop2 <- loan_new  |>
  # Group by "group"
  group_by(loan_cat) |>
  count(homeownership) |>
  # Create new variable, prop, using mutate
  mutate(prop = n/sum(n)) |>
  filter(homeownership == "rent")
prop2

```

*We can see that renters had higher proportions of loans below the average loan amount.*
