---
title: "Chapter 18 - Chi Square"
author: "R Saidi"
format: docx
editor: visual
---

## Load the libraries and data

```{r}
#| message: false
library(tidyverse)
library(openintro)
library(tidymodels)
data("gss")
```

## A question in two variables

Does a college degree or not have an association with political party affiliation?

Is having a college degree or not independent of political party affiliation?

HO: Having a college degree or not is independent of political party affiliation
Ha: Having a college degree or not is NOT independent of political party affiliation

## Chi Square

When we are looking at a two-way table, we can explore the $\chi^2$ distribution

## Start with a bar plot

```{r}
# Visualize distribution
ggplot(data = gss, aes(x = partyid, fill = college)) +
  # Add bar layer of proportions
  geom_bar(position = "fill")
```

*The education proportions for each party look relatively similar.*

## Remove position = "fill"

```{r}
gss |>
  ggplot(aes(x = partyid, fill = college)) +
  # Add bar layer of proportions
  geom_bar()
```

## base R table

```{r}
obs_table <- table(gss$college, gss$partyid)
obs_table
```

```{r}
# From previous step
obs <- gss |>
  select(college, partyid) |>
  tibble::as_tibble() |>
  table()
obs
```

## What is the DK?

DK seems to be some anomaly that needs to be removed.

```{r}
new_df <- gss |>
  filter(partyid != "DK")
unique(new_df$partyid)
```

```{r}
gss$partyid <- as.character(gss$partyid) |> 
  trimws() |> 
  as.factor() 
```

```{r}
# Now check that DK is removed
unique(gss$partyid)
```

# Convert table back to tidy df

```{r}
obs |>
  # Tidy the table
  tidy() |>
  # Expand out the counts
  uncount(n)
```

```{r}
gss |>
  group_by(partyid)|>
  count(college)
```

## Base R approach

```{r}
chisq.test(gss$college, gss$partyid)$expected
```



## Perform a chi-squre hypothesis test

1.  Calculate the observed Chi-Square statistic

```{r}
# calculate the observed statistic
observed_indep_statistic <- gss |>
  specify(partyid ~ college) |>   #y ~ x
  hypothesize(null = "independence") |>
  calculate(stat = "Chisq")
observed_indep_statistic
```

```{r}
# Create one permuted data set
perm_null <- gss |>
  specify(partyid ~ college) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "Chisq")
perm_null
```

```{r}
perm_null |>
  visualize() +
  shade_p_value(observed_indep_statistic,
    direction = "greater"
  )
```

## calculate the p-value under tidymodels

```{r}
pvalue <- get_p_value(perm_null, observed_indep_statistic, direction = "two.sided")
pvalue
```

## Compare to base R

```{r}
# Compute chi-squared stat
chisq.test(gss$partyid, gss$college)
```

*With a p-value of 0.3608, fail to reject the null. There is no compelling evidence that there is an association between having a college degree or not and the political party affiliation.*

or

*Whether or not a person has a college degree is independent of party affiliation.*

## Fisher's Exact Test

When basic assumptions for Chi-Square test (expected cell counts \>= 5) are violated, we can try using the comparable non-parametric Fisher's Exact Test.

p-value = (rowtot C success)\*(rowtot C failure) / (samplesize C truetotal)

## Example Below


```{r}
# fisher.test(v1, v2)
fisher.test(gss$college, gss$partyid)
```


