---
title: "Chapters 20 interactive notes"
author: "R Saidi"
format: docx
editor: visual
---

# Chapter 20 - Difference of Two Groups' Means

## load the libraries and dataset

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(openintro)
library(tidymodels)
data(births14)
data(stem_cell)
```

This dataset is a sample of 1000 observations.

## Check basic assumptions

Use a side-by-side boxplot to compare the two groups' distributions

```{r}
ggplot(births14, aes(x=habit, y=weight, fill = habit))+
  geom_boxplot()+
  geom_jitter(alpha = .1)
```

Notice there are some na values. Let's remove those.

```{r}
births14_a <- births14 |>
  filter(!is.na(habit))
unique(births14_a$habit)
```

```{r}
ggplot(births14_a, aes(x=habit, y=weight, fill = habit))+
  geom_boxplot()+
  geom_jitter(alpha = .2)
```

## Check sample sizes are sufficiently large

this will check the basic assumptions for t-test

```{r}
births_table <- births14_a |>
  group_by(habit) |>
  count()
births_table
```

## Now replot, but use density plots

```{r}
ggplot(births14_a, aes(weight, fill = habit))+
  geom_density(alpha = .2)
```

*We can see the density plots are skewed left, but the sample size is 981, so it passes the CLT)*

## t.test for difference of means in base R

Use a t.test for difference of means to compare mean birth weights of babies born to smoking mothers versus non-smoking mothers.

Use the syntax: t.test(y ~ x)

```{r}
t.test(births14_a$weight ~ births14_a$habit)
```

## Conclusion

Reject the null. There is very strong evidence that the mean birthweight for newborns is different between smoking and non-smoking mothers. 


**t = 3.8166, df = 131.31, p-value = 0.0002075**

## Alternate code using t_test

```{r}
# use package: infer (in tidymodels)
t_test <- births14_a |> 
  t_test(response = weight,
         explanatory = habit,
         null = 0, 
         alternative = "two-sided", 
         conf_int = TRUE)
t_test
```

## Alternate code using randomization

use specify, hypothesize, generate, calculate

### Calculate the difference in observed means

Use specify and calculate

```{r}
# Calculate observed difference in means
diff_mean_obs <- births14_a |>
  # Specify the response and explanatory variables
  specify(weight ~ habit) |>           # syntax is y ~ x
  calculate(stat = "diff in means", order = c("nonsmoker", "smoker")) 
diff_mean_obs
```

```{r}
diff_ht_mean <- births14_a |>
  specify(weight ~ habit) |>                    # syntax is y ~ x 
  hypothesize(null = "independence") |>         # set the null
  generate(reps = 1000, type = "permute") |>    # shuffle 1000 times
  calculate(stat = "diff in means", order = c("nonsmoker", "smoker")) 
# Specify to calculate a difference in means and what order of subtraction to use
head(diff_ht_mean)
```

## View a histogram of randomized distribution under the null

```{r}
ggplot(diff_ht_mean, aes(x=stat))+
  geom_histogram()+
  geom_vline(xintercept = pull(diff_mean_obs), color = "blue")
```

## Calculate the p-value

```{r}
diff_ht_mean |>
  get_p_value(obs_stat = diff_mean_obs, direction = "two-sided") 
```

## Conclusion

Reject the null. There is very strong evidence that there is a difference in mean birth weights for babies born to smoking and non-smoking mothers.

```{r}
diff_ht_mean |>
  visualize() +
  shade_p_value(diff_mean_obs, direction = "two-sided")
```

## Example using stem_cell data

Does treatment using embryonic stem cells (ESCs) help improve heart function following a heart attack? Table 20.2 contains summary statistics for an experiment to test ESCs in sheep that had a heart attack. Each of these sheep was randomly assigned to the ESC or control group, and the change in their hearts' pumping capacity was measured in the study.

Our goal will be to identify a 95% confidence interval for the effect of ESCs on the change in heart pumping capacity relative to the control group.

| Group   |  n  | Mean  |  SD  |
|:--------|:---:|:-----:|:----:|
| ESC     |  9  | 3.50  | 5.17 |
| Control |  9  | -4.33 | 2.76 |

## When you only have summary statistics to work with

Use the function tsum.test(x-bar1, s1, n1, x-bar2, s2, n2) in the package BSDA.

```{r}
library(BSDA)
tsum.test(3.5, 5.17, 9, -4.33, 2.76, 9)
```

## Bootstrap code for difference in 2 group means

Comparing the mean birthweight between newborns of non-smoking and smoking mothers.

```{r}
diff_mean_ci <- births14_a |>
  # Specify weight vs. habit
  specify(weight ~ habit) |>
  # Generate 1500 bootstrap replicates
  generate(reps = 1000, type = "bootstrap") |>
  # Calculate the difference in means, nonsmoker then smoker
  calculate(stat = "diff in means", order = c("nonsmoker", "smoker"))
diff_mean_ci
```

## Now calculate the 95% CI

```{r}
# Calculate the 95% CI via percentile method
diff_mean_ci |>
  get_ci(level = 0.95)
```

### Conclusion:

Zero is not included. We are 95% confident that the true difference in mean weight for babies is between 0.31 and 0.90 lbs higher for nonsmoking mothers versus smoking mothers.

# What to do when the basic assumptions fail for a t.test

When the sample sizes for each group are small AND the distributions are not bell-shaped.

Use the raw data from the stem_cell dataset from openintro.org

```{r}
stem <- stem_cell |>
  mutate(diff = after - before)
head(stem)
```

Side-by-side boxplots

```{r}
ggplot(stem, aes(x=trmt, y=diff))+
  geom_boxplot()+
  geom_jitter(alpha = .5)
```

histograms/density plots

```{r}
ggplot(stem, aes(diff, fill = trmt))+
  geom_density(alpha = .2)
```

## When checking the basic assumptions, check the distribution AND the sample sizes

```{r}
stem |>
  group_by(trmt)|>
  count()
```

*We can see the sample sizes are VERY small. We need to use a non-parametric approach.*

## Non-parametric approach (not computing the means)

Wilcox Mann Whitney Test

Null: There is no difference in heart pumping capacity between control and treatment groups.

Alternative: There is a shift in distributions between the control and treatment groups for heart pumping capacity.

```{r}
wilcox.test(stem$diff ~ stem$trmt,  conf.int = TRUE, conf.level = 0.95,) #y~x
```

*Conclusion: Because the p-value is very small, there is a right shift in the difference in heart pumping capacity between treatment for the ESC group over the control group.*
