---
title: "Math217_ch5_notes"
format: docx
editor: visual
---

# Working with single numerical variables

## Numerical variables: the distribution of shape, center, and spread

When you want to know more about a function, use a question mark with the function

Example: learn more about the function, boxplot

```{r}
# ?boxplot   (delete the # to run ?boxplot)
```


## I. Shape

1. Dotplots (not necessarily the best way to show distributions)


```{r}
# load the vector of data for number of segments

segment <- c(23, 30, 54, 28, 31, 29, 34, 35, 30, 27, 21, 43, 51, 35, 51, 49, 35, 24, 26, 29, 21, 29, 37, 27, 28, 33, 33, 23, 37, 27, 40, 48, 41, 20, 30, 57)
```

```{r}
# stripchart is a base-R dotplot
stripchart(segment)
```

2. Histograms are better at displaying a numerical variable's shape

```{r}
hist(segment, col = "lightblue")
```

*We can see that the distribution of this variable is right-skewed (the tail is to the right)*

3. Boxplots show OTHER information about the distribution, based on the 5-number summary

```{r}
# 5-number summary + mean
summary(segment)
```

```{r}
# now create the boxplot using these values
boxplot(segment, col = "lightpink")
```

* Notice that there is one upper outlier, which is 57, based on the max value from the summary output.

* The median is 30.5.

* The IQR (Q3-Q1 = 37.75-27 = 10.75) is the middle box

```{r}
# Rotate your boxplot horizontally and add a notch at the median.
# Add a title (main =) and x-axis label (xlab =)
boxplot(segment, col = "lightpink", notch = TRUE, horizontal = TRUE, main = "Notched boxplot of number of segments", xlab = "Number of Segments")
```

## II. Center

The two measure of center for a numerical variable are:
**mean** and **median**

```{r}
mean(segment)
median(segment)
```

*When the mean and median are essentially the same value, that indicates a bell shape. When mean is greater than median (as it is in this case), the distribution is right-skewed, since the mean is dragged towards the tail.*


## III. Spread

Spread describes how wide the values are across a distribution. We can use the following:

* range =  max - min

* IQR =  Q3 - Q1

* standard deviation - this has a complicated formula (see notes). The function in R is "sd"

```{r}
IQR(segment)
range(segment)
sd(segment)
```


## Your Turn!!!

### Use the following data: sodium

Sodium is a vector containing the grams of sodium in 20 cereal boxes.

```{r}
sodium <- c(0,	340,	70,	140,	200,	180,	210,	150,	100,	130,	140,	180,	190,	160,	290,	50,	220,	180,	200,	210)
```


## Create a chunk for each question below and use appropriate code to answer.

1. Create a dotplot, histogram, and boxplot of the distribution of sodium.

```{r}
# dotplot
```

```{r}
# histogram
```

```{r}
# boxplot
```


2. Find the 5-number summary plus mean for sodium.

```{r}

```

3. What is the IQR and sd?

```{r}

```

4. Make a histogram or boxplot, but add a title, axis labels, and change the color.

```{r}

```


