---
title: "Chapter 18 - Chi Square"
author: "R Saidi"
format: docx
editor: visual
---

## Load the libraries and data

```{r}
#| message: false
library(tidyverse)
library(openintro)
library(tidymodels)
data("gss")
```

## A question in two variables

Does level of education have an association with political party affiliation?

## Chi Square

When we are looking at a two-way table, we can explore the $\chi^2$ distribution

## Start with a bar plot

```{r}
# Visualize distribution
gss |>
  ggplot(aes(x = partyid, fill = college)) +
  # Add bar layer of proportions
  geom_bar(position = "fill")
```

*The education proportions for each party look relatively similar.*

## Remove position = "fill"

```{r}
gss |>
  ggplot(aes(x = partyid, fill = college)) +
  # Add bar layer of proportions
  geom_bar()
```

## base R table

```{r}
obs_table <- table(gss$college, gss$partyid)
obs_table
```


```{r}
# From previous step
obs <- gss |>
  select(college, partyid) |>
  tibble::as_tibble() |>
  table()
obs
```

## What is the DK? 

DK seems to be some anomaly that needs to be removed. 

```{r}
gss$partyid <- as.character(gss$partyid) |> 
  trimws() |> 
  as.factor() 
```


```{r}
# Now check that DK is removed
unique(gss$partyid)
```

# Convert table back to tidy df

```{r}
obs |>
  # Tidy the table
  tidy() |>
  # Expand out the counts
  uncount(n)
```

## Perform a chi-squre hypothesis test

```{r}
# Create one permuted data set
perm_1 <- gss |>
  # Specify the variables of interest
  specify(college ~ partyid) |>
  # Set up the null
  hypothesize(null = "independence") |>
  # Generate a single permuted data set
  generate(reps = 1, type = "permute")
perm_1
```

```{r}
# Visualize permuted data
ggplot(perm_1, aes(x = partyid, fill = college)) +
  # Add bar layer
  geom_bar()
```

```{r}
# Compute chi-squared stat
chisq.test(perm_1$partyid, perm_1$college)
```

*With a p-value of 0.3608, there is no compelling evidence that there is an association between having a college degree or not and the political party affiliation.*


## Fisher's Exact Test

When basic assumptions for Chi-Square test (expected cell counts >= 5) are violated, we can try using the comparable non-parametric Fisher's Exact Test.


(rowtot C success)*(rowtot C failure) / (samplesize C truetotal)

## Example Below

      true  false  total
     
yes     4     3      7
no     13     8     21
total  17    11     28
      
FE  =  ((7 C 4)  * (21 C 13)) / (28 C 17)

```{r}
# fisher.test(v1, v2)
fisher.test(gss$college, gss$partyid)
```

