---
title: "Chapter 16 Interactive Notes"
author: "R Saidi"
date: "2023-10-22"
output:
  word_document: default
  html_document: default
---


## Base R approach to looking at complication rates for liver donor surgeries

```{r}
prop.test(x = 3, n = 62, conf.level = .95, p = .10)
```



## Load the libraries and data

```{r}
#| message: false
library(tidyverse)
library(openintro)
library(tidymodels)
data("resume")
```

## Rename the factor levels for callbacks

```{r}
resume1 <- resume |>
  mutate(called = ifelse(received_callback == 1, "yes", "no"))
```

## View the counts for applicants receiving callbacks

```{r}
ggplot(resume1, aes(x = called)) +
  geom_bar()
```

## calculate exact proportion of the sample that responded this way: p-hat

```{r}
p_hat_call <- resume1 |>
  summarize(prop_call = mean(called == "yes")) |>
  pull()
p_hat_call
```

*The proportion who received a callback in the dataset was 0.080.*

## Test Basic Assumptions

* observations are independent
* np_hat >= 10 n(1-p_hat) >= 10  (Success-Failure Condition)

```{r}
resume1 |>
  group_by(called)|>
  count()
```



### The 95% confidence interval can be calculated as the sample proportion plus or minus two standard errors of the sample proportion

The bootstrap is done with the function: **specify()**

### We do this many times to create many bootstrap replicate data sets.

Do this with the function **generate()**

### Next, for each replicate, we calculate the sample statistic, in this case: the proportion of respondents that said "yes" to receiving callbacks.

Do this with the function **calculate()**

```{r}
boot_dist_call <- resume1 |>
  specify(response = called, success = "yes") |> # single group
  generate(reps = 1000, type = "bootstrap") |>
  calculate(stat = "prop")
boot_dist_call
```

## Plot the density curve of this distribution

```{r}
ggplot(boot_dist_call, aes(x = stat)) +
  geom_density()
```

The standard deviation of the stat variable in this data frame (the bootstrap distribution) is the bootstrap standard error and it can be calculated using the summarize() function.

```{r}
SE_call <- boot_dist_call |>
  summarize(se = sd(stat)) |>
  pull()
SE_call
```


```{r}
# manual/formulaic computation for se

se <- sqrt((p_hat_call*(1-p_hat_call))/62)
se
```

## margin of error

z* - cutoff for the 95% confidence interval 1.96
se - either formulaic or bootstrap value

```{r}
me <- 1.96*se
me
```



## We can use this value, along with our point estimate, to roughly calculate a 95% confidence interval:

$\hat{p} \pm z^*se$

```{r}
c(p_hat_call - 1.96 * SE_call, p_hat_call + 1.96 * SE_call)
```

*We are 95% confident that the true proportion of applicants receiving callbacks is between 7.28% and 8.81%.*

## A more efficient way to get the CI

Use get_ci


```{r}
ci <- get_ci(boot_dist_call, level=.95)
ci
```

*We are 95% confident that the true proportion of applicants receiving callbacks is between 7.33% and 8.77%.*


## The normal distribution for confidence interval

Another option for calculating the CI is by estimating it by using the Normal Distribution (the bell curve)

**If**

1.  observations are independent
2.  n is large (S-F condition is met)

**Then**

$\hat{p}$ follows a normal distribution

## Steps

### Calculate proportion receiving callback

```{r}
n <- nrow(resume1)
p_hat_call <- resume1|>
  summarize(prop_call = mean(called == "yes")) |>
  pull()
p_hat_call
```

### Check conditions (both should be \>= 10)

```{r}
n * p_hat_call
n * (1 - p_hat_call)
```

### Calculate SE

```{r}
se_call_approx <- sqrt(p_hat_call * (1 - p_hat_call) / n)
```

### Calculate

$z^*$

```{r}
z_star <- qnorm(.975, m=0, sd = 1)
```

### Form 95% CI

```{r}
c(p_hat_call - z_star * se_call_approx, p_hat_call + z_star * se_call_approx)
```


## Hypothesis Test

Use the General Social Survey (gss) data

```{r}
data("gss")
head(gss)
```

### Do 1/3 of Americans have a college degree?

Use a hypothesis test to study a question on the gss. Respondents were asked if have a college degree. You can look at the distribution of answers by forming a bar chart. You see that of the 500 respondents, about 180 have a degree.

```{r}
ggplot(gss, aes(x=college)) +
  geom_bar()
```

## State the null and alternative hypotheses

$H_o: p = 1/3$
Ho: The proportion of US adults who have a college degree is 1/3.

$H_a: p \neq 1/3$
Ha: The proportion of US adults who have a college degree is different from 1/3.


## Check Success-Failure Condition

The difference with S-F condition for hypothesis testing is to use the population parameter instead of the p-hat value.

np >=10
n(1-p)>=10

```{r}
s <- (1/3)*500
s
f <- (2/3)*500
f
```

*n=Notice the s-f condition has been met for this hypothesis test.*

## Calculate the sample proportion

```{r}
p_hat <- gss |>
  summarize(mean(college == "degree")) |>
  pull()
p_hat
```

p-hat = 0.348



## Now perform the hypothesis test

```{r}
null_dist <- gss |>
  specify(response = college, success = "degree") |>
  hypothesize(null = "point", p = 1/3) |>   #null = "point" single mean or proportion
  generate(reps = 1000, type = "draw") |>   #type = "draw" single proportion
  calculate(stat = "prop")
null_dist
```

## Create a density curve of the null distribution

Add a vertical line (geom_vline) to indicate where p-hat lies on the curve.

```{r}
ggplot(null_dist, aes(x = stat)) +
  geom_density() +
  geom_vline(xintercept=1/3, color = "red", linetype = "dashed" ) + # center for hypothesis test
  geom_text(x = 1/3, y = 3, color = "red", label = "null")+
  geom_vline(xintercept = p_hat, color = "cyan")
```

## Calculate a two-tailed p-value (multiply by two)


```{r}
pvalue2 <- get_p_value(null_dist, p_hat, direction = "two-sided")
pvalue2
```


*The p-value is very large, at 0.534. We fail to reject the null. There is no evidence that the proportion of US adults with a college degree is different than 1/3 the population.*



