---
title: "Logistic Regression"
author: "Rachel Saidi"
date: "5/7/2020"
format: docx
---

## Load the IST Dataset (International Stroke Trial Database)

<https://datashare.is.ed.ac.uk/handle/10283/128>

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
setwd("C:/Users/rsaidi/Dropbox/Rachel/MontColl/Datasets/Datasets")
IST <- read_csv("IST.csv")
IST_vars <- read_csv("IST variables.csv")
# this csv defines the variables
```

## Create a table of survival based on sex

Note: DIED == 0 means died and DIED == 1 means alive

```{r}
table(IST$SEX, IST$DIED)
table(IST$SEX, IST$DIED) |> 
  prop.table() # convert the table to proportions
```

## Just a bit of cleaning

```{r}
IST$RXASP <- as.factor(IST$RXASP)
IST$RXHEP <- as.factor(IST$RXHEP)
IST1 <- IST |>
  drop_na(c(AGE, SEX, RXASP, RXHEP, DDIAGUN, DDIAGHA, DDIAGISC, RDEF1, RDEF2, RDEF3, RDEF4, RDEF5, RDEF6, RDEF7, RDEF8))
```

## Machine Learning

```{r}
set.seed(1234)
IST_samp <- sample_n(IST1, size = 300, replace = TRUE)
res_splits<- initial_split(IST_samp, strata = DIED)

died_train <- training(res_splits) #default 75% 
died_test  <- testing(res_splits) # default 25%
```

## Calculate training set proportions by is_alive

```{r}
died_train |>
  group_by(DIED) |>
  tally() |>
  mutate(prop = n/sum(n))
```

## Fit a logistic regression full model

Note there are 112 variables. We will only select the most important ones for the full model, knowing this is a biased approach.

```{r}
# fit1 <- glm(DIED ~ AGE + RXHEP + RXASP + DSTER + HOSPNUM + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA + DDIAGUN + FAP,  family = binomial(), data = died_train)  #generalized linear model
# summary(fit1)
# one of the variables is causing a problem
```

## The problem is the FAP variable, so remove it

```{r}
fit2 <- glm(DIED ~ AGE + RXHEP + RXASP + DSTER + HOSPNUM + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA + DDIAGUN,  family = binomial(), data = died_train)  #generalized linear model
summary(fit2)
```

## Remove HOSPNUM

```{r}
fit3 <- glm(DIED ~ AGE + RXHEP + RXASP + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA + DDIAGUN,  family = binomial(), data = died_train)  #generalized linear model
summary(fit3)
```

## Remove DDIAGUN

```{r}
fit4 <- glm(DIED ~ AGE + RXHEP + RXASP + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA,  family = binomial(), data = died_train)  #generalized linear model
summary(fit4)
```

## Remove RXHEP (heparin prescription)

```{r}
fit5 <- glm(DIED ~ AGE + RXASP + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA, family = binomial(), data = died_train)  #generalized linear model
summary(fit5)
```

## Remove RXASP (aspirin prescription)

```{r}
fit6 <- glm(DIED ~ AGE + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA, family = binomial(), data = died_train)  #generalized linear model
summary(fit6)
```

## Remove DSTER (steroid prescription)

```{r}
fit7 <- glm(DIED ~ AGE + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA, family = binomial(), data = died_train)  #generalized linear model
summary(fit7)
```

## Remove RDEF4 and RDEF8

```{r}
fit8 <- glm(DIED ~ AGE + RATRIAL + RDEF5 + RDEF6 + RDEF7+ DDIAGISC + DDIAGHA, family = binomial(), data = died_train)  #generalized linear model
summary(fit8)
```

## Remove RATRIAL and RDEF7

```{r}
fit9 <- glm(DIED ~ AGE + RDEF5 + RDEF6 + DDIAGISC + DDIAGHA, family = binomial(), data = died_train)  #generalized linear model
summary(fit9)
```

## Remove RDEF6

```{r}
fit10 <- glm(DIED ~ AGE + RDEF5 + DDIAGISC + DDIAGHA, family = binomial(), data = died_train)  #generalized linear model
summary(fit10)
```

### Result of the final model

$log\frac{DIED}{(1-DIED)} = 0.198 + 0.024(Age) - 2.199(Hemianopia_{no}) - 1.365(Hemianopia_{yes}) -1.799(Ischaemic stroke_{yes}) - 1.663(Haemorrhagic stroke_{yes})$


With AIC: 215.14

## Summary of Feature Selection

| Model | Variables | AIC |
|---------------------|---------------------------------|------------------|
| fit1 | AGE + RXHEP + RXASP + DSTER + HOSPNUM + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA + DDIAGUN + FAP | model failed |
| fit2 | AGE + RXHEP + RXASP + DSTER + HOSPNUM + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA + DDIAGUN | 200.4 |
| fit3 | AGE + RXHEP + RXASP + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA + DDIAGUN | 198.2 |
| fit4 | AGE + RXHEP + RXASP + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA | 196.43 |
| fit5 | AGE + RXASP + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA | 193.54 |
| fit6 | AGE + DSTER + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA | 191.55 |
| fit7 | AGE + RATRIAL + RDEF4 + RDEF5 + RDEF6 + RDEF7 + RDEF8 + DDIAGISC + DDIAGHA | 189.97 |
| fit8 | AGE + RATRIAL + RDEF5 + RDEF6 + RDEF7+ DDIAGISC + DDIAGHA | 186.96 |
| fit9 | AGE + RDEF5 + RDEF6 + DDIAGISC + DDIAGHA | \*212.47 |
| fit10 | AGE + RDEF5 + DDIAGISC + DDIAGHA | \*215.14 |


## Sample Interpretation

```{r}

age_50 <- data.frame(AGE = 50, RDEF5 = "Y", DDIAGISC = "Y", DDIAGHA = "N")

augment(fit10, newdata = age_50, type.predict = "response")
augment(fit10, newdata = age_50, type.predict = "response")
exp(0.15047)
p = 1.16238/2.16238
p
```

This means the odds of dying for a 50 year old with Hemianopia and ischaemic stroke:

is:

(p/1-p) = exp(0.15)

(p/1-p) = 1.16238 \# Odds of dying for 55 a 55 year old in this population

p = 1.16238/2.16238 = 0.5375466 \# Probability of dying for 50 year old in this population is 0.537 or 53.7%.

## Create a confusion matrix manually

- is_alive is the observed value
- alive_hat is the predicted value

```{r}
fit_plus <- augment(fit10, type.predict = "response") |>
  mutate(died_hat = round(.fitted))

fit_plus |>
  select(DIED, died_hat) |>
  table()
```

## Manually calculate accuracy

```{r}
accuracy_new <- (156 + 27)/(156 + 27 + 27 + 15)
accuracy_new
```

## The current accuracy of this training model is 81.3%

## Illustrate the predicted probability

Illustrate how the predicted probability of dying varies in our simple logistic regression model with respect to a person's age. Notice the sigmoid curve (in purple) - this is the shape of a logistic curve.

```{r}
ggplot(died_train, aes(x = AGE, y = DIED)) + 
  geom_count(
    position = position_jitter(width = 0, height = 0.05), 
    alpha = 0.5 ) + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              color = "purple", lty = 2, se = FALSE) + 
  geom_hline(aes(yintercept = 0.5), linetype = 3) 
```

## Note the sigmoid curve increases as age increases

The probability of dying increases with age.

## Use tidymodels to build the model

Create a categorical variable `died_class` that is Y/N for 1/0 for both the testing and training datasets

```{r}
died_train1 <- died_train |>
  mutate(died_class = ifelse(DIED == 0, "No", "Yes"), .after = DIED)
died_train1$died_class <- as.factor(died_train1$died_class)

died_test1 <- died_test |>
  mutate(died_class = ifelse(DIED == 0, "No", "Yes"), .after = DIED)
died_test1$died_class <- as.factor(died_test1$died_class)

logit_fit_train <- 
  logistic_reg(mode = "classification") |>
  set_engine(engine = "glm") |> 
  fit(died_class ~ AGE + RDEF5 + DDIAGISC + DDIAGHA, family = binomial(), data = died_train1)
logit_fit_train
```

## Model Prediction (tidymodel)

Test Data Class Prediction

The very next step is to generate the test predictions that we could use for model evaluation. To generate the class prediction (pos/ neg) we can use the predict function and supply the trained model object, test dataset and the type which is here “class” as we want the class prediction, not probabilities.

# To get predicted values we use the function **predict**.

```{r}
pred_class <- predict(logit_fit_train,
                      new_data = died_test1,
                      type = "class")

pred_prob <- predict(logit_fit_train,
                      new_data = died_test1,
                      type = "prob")

joined_pred_results <- died_test1 |>
   select(died_class, DIED) |>
   bind_cols(pred_class, pred_prob)
```

## Model performance (tidymodel)

**yardstick** is a package to get metrics on model performance. By default on classification problem yardstick::metrics returns accuracy and kappa.

<https://yardstick.tidymodels.org/>

```{r}
library(yardstick)
metrics(joined_pred_results, truth = died_class, estimate = .pred_class)
```

*Accuracy is about 73.3% and kappa is 0.17. A kappa of 1 means perfect agreement with the two categorical classifications, and a kappa of 0 means no agreement (no better than random assignment). This is relatively low but not zero.*

## Sensitivity and Specificity (tidymodel)

```{r}
sens <- sens(joined_pred_results, truth = died_class, estimate = .pred_class)
spec <- spec(joined_pred_results, truth = died_class, estimate = .pred_class)
rbind(sens, spec)
```

*The sensitivity is high at 87.7% but specificity is very low at 27.8%.*

## VIF - variance inflation factor (base R model)

Values between 5-10 indicate collinearity

```{r}
#install.packages("car") for base R model
library(car)
vif(fit10)

# vif for tidymodel
base_fit <- extract_fit_engine(logit_fit_train)
vif(base_fit)
```

\*All values are low between 1 and 2.

## ROC Curve - Receiver Operator Characteristics

Now, we will use the ROC curve along with its corresponding AUC (area under the curve) for the testing set.

## ROC - Receiver Operating Characteristics

Receiver Operating Characteristics Curve traces the percentage of true positives accurately predicted by a given logit model as the prediction probability cutoff is lowered from 1 to 0. For a good model, as the cutoff is lowered, it should mark more of actual 1's as positives and lesser of actual 0's as 1's. So for a good model, the curve should rise steeply, indicating that the TPR (Y-Axis) increases faster than the FPR (X-Axis) as the cutoff score decreases.

## AUC - Area under the curve

The greater the area under the ROC curve, better the predictive ability of the model

```{r}
roc_auc(bind_cols(died_test1, pred_prob), died_class, .pred_No)
```

*Area under the curve is 74.4%*

```{r}
roc_data <- roc_curve(bind_cols(died_test1, pred_prob), truth = died_class, .pred_No) 

roc_data |>  
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_smooth() +
  geom_abline(lty = 3) + 
  coord_equal()+
  ggtitle("ROC Curve")+
  theme_bw()+
  geom_text(x=0.2, y=.8, label="AUC = 74.4%", color ="purple")
```



## Final model

$log\frac{DIED}{(1-DIED)} = 0.198 + 0.024(Age) - 2.199(Hemianopia_{no}) - 1.365(Hemianopia_{yes}) -1.799(Ischaemic stroke_{yes}) - 1.663(Haemorrhagic stroke_{yes})$
